Arrows for Parallel Computation

### 2019-12-27 paper/content/arxiv/ms.tex
reviewer's comments (1)

We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner -- we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel implementations at will. The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}). Further implementations, based on HdpH or a Frege implementation (on the Java Virtual Machine), are viable, too.

@@

\ptcomment{The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations. A range of approaches have been explored, including data parallelism (Chakravarty et al., 2007; Keller et al.,
2010), GPU-based approaches (Mainland & Morrisett, 2010; Svensson, 2011), software
transactional memory (Harris et al., 2005; Perfumo et al., 2008). The Haskell–GPU bridge
Accelerate (Chakravarty et al., 2011; Clifton-Everest et al., 2014; McDonell et al., 2015)
deserves a special mention. <<Why does it deserve special mention?>>  Accelerate is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.

@@

Other task parallel Haskells, related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad, include the following.  
HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of \ensuremath{\Conid{Par}} Monad to
heterogeneous clusters. LVish (Kuper et al., 2014) is a communication-centred extension
of \ensuremath{\Conid{Par}} Monad.}

@@

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, in essence they are a generalised function arrow~\ensuremath{\to }. \citet{Hughes2005} presents a tutorial on Arrows. \citet{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} develop theoretical background of Arrows. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular Arrows do and thus enable optimisations. Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of list of Arrows. \citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.

\paragraph{Arrows in other languages.}
Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined. Basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM). However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}}, 
a new approach to Haskell on the JVM.

@@

While we could have based our DSL on Profunctors as well, we chose Arrows\mbcomment{cite missing?} for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability. However, they are a promising candidate for future improvements of our DSL. Some Profunctors, especially ones supporting a composition operation, choice, and looping, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
	\subsection{Short introduction to parallel Haskells}
\label{sec:parallelHaskells}
\label{sec:parEvalNIntro}
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel or \ensuremath{\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{parEvalN}
	\caption{Schematic illustration of \ensuremath{\Varid{parEvalN}}. A list of inputs is transformed by different functions in parallel.}
	\label{fig:parEvalN}
\end{figure}

In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow-based parallel Haskell.

There exist several parallel Haskells already. Among the most important are probably GpH \citep[based on \ensuremath{\Varid{par}} and \ensuremath{\Varid{pseq}} \enquote{hints},][]{Trinder1996,Trinder1998a}, the \ensuremath{\Conid{Par}} Monad \citep[a monad for deterministic parallelism,][]{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \citep[a parallel Haskell for distributed memory,][]{eden,Loogen2012}, HdpH \citep[a Template Haskell-based parallel Haskell for distributed memory,][]{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish \citep[a \ensuremath{\Conid{Par}} extension with focus on communication,][]{Kuper:2014:TPE:2666356.2594312}.

As the goal of this paper is not to re-implement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface. For this paper we chose GpH for its simplicity, the \ensuremath{\Conid{Par}} Monad to represent a monadic DSL, and Eden as a distributed parallel Haskell.

LVish and HdpH were not chosen as the former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper, while the latter (at least in its current form) does not comply with our representation of parallelism due to its heavy reliance on Template Haskell.

We will now go into some detail on GpH, the \ensuremath{\Conid{Par}} Monad and Eden, and also give their respective implementations of the non-Arrow version of \ensuremath{\Varid{parEvalN}}.

%Before we go into detail on how we can use this idea of parallelism
%for parallel Arrows, as a short introduction to parallelism in
%Haskell we will now implement |parEvalN| with several different
%parallel Haskells.

@@

\paragraph{Eden TraceViewer.}
\label{sec:edentv}
To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}. In the next sections we will present some \emph{trace visualisations}, the post-mortem process diagrams of Eden processes and their activity.

@@

\subsubsection{\ensuremath{\Conid{Par}} Monad}
\olcomment{introduce a newcommand for par-monad, "Arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
As for GpH we can easily lift the definition of \ensuremath{\Varid{parEvalN}} for the \ensuremath{\Conid{Par}} Monad to Arrows in Fig.~\ref{fig:ArrowParallelParMonad}. To start off, we define the \ensuremath{\Conid{Strategy}\;\Varid{a}} and \ensuremath{\Conid{Conf}\;\Varid{a}} type so we can have a configurable instance of ArrowParallel:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{Strategy}\;\Varid{a}\mathrel{=}\Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{Conf}\;\Varid{a}\mathrel{=}\Conid{Conf}\;(\Conid{Strategy}\;\Varid{a}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Now we can once again define our \ensuremath{\Conid{ArrowParallel}} instance as follows: First, we convert our Arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{evalN}\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{strat})\;\Varid{fs})} into an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu (\Conid{Par}\;(\Conid{IVar}\;\Varid{b}))\mskip1.5mu]} that yields composable computations in the \ensuremath{\Conid{Par}} monad. By combining the result of this Arrow with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}, we get an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu])}. Then, in order to fetch the results of the different threads, we map over the \ensuremath{\Conid{IVar}}s inside the \ensuremath{\Conid{Par}} Monad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})} -- our intermediary Arrow is of type \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu])}. Finally, we execute the computation \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} by composing with \ensuremath{\Varid{arr}\;\Varid{runPar}} and get the final Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{parEvalN}\;(\Conid{Conf}\;\Varid{strat})\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{evalN}\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{strat})\;\Varid{fs})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{\Conid{Control}.\Conid{Monad}.\Conid{Par}.get})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{runPar}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{\ensuremath{\Conid{Par}} Monad \ensuremath{\Conid{ArrowParallel}} instance.}
\label{fig:ArrowParallelParMonad}
\end{figure}

@@

\subsubsection{Default configuration instances}
While the configurability in the instances of the \ensuremath{\Conid{ArrowParallel}} instances above is nice, users probably would like to have proper default configurations for many parallel programs as well. These can also easily be defined as we can see by the example of the default implementation of \ensuremath{\Conid{ArrowParallel}} for the GpH:

@@

This is only a problem in distributed memory (in the scope of this paper) and we should allow nodes to communicate directly with each other. Eden already provides \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
But as we want code with our DSL to be implementation agnostic, we have to wrap this concept. We do this with the \ensuremath{\Conid{Future}} type class (Fig.~\ref{fig:future}). We require a \ensuremath{\Varid{conf}} parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the \ensuremath{\Conid{ArrowParallel}} type class earlier.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{Future}\;\Varid{fut}\;\Varid{a}\;\Varid{conf}\mid \Varid{a}\;\Varid{conf}\to \Varid{fut}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{put}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{conf}\to \Varid{arr}\;\Varid{a}\;(\Varid{fut}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{get}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{conf}\to \Varid{arr}\;(\Varid{fut}\;\Varid{a})\;\Varid{a}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The \ensuremath{\Conid{Future}} type class.}
\label{fig:future}
\end{figure}
Since \ensuremath{\Conid{RD}} is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows. %This is due to the same reason we had to introduce a wrapper for |Strategy a| in the GpH Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
Technical details are in Appendix, in Section~\ref{app:omitted}.

@@

Next, we zip the resulting \ensuremath{([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])} to \ensuremath{[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})}. We then feed this into our parallel Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]\;[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} obtained by transforming our input Arrow \ensuremath{\Varid{f}\mathbin{::}\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})} into \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})} before \ensuremath{\Varid{repeat}}ing and lifting it with \ensuremath{\Varid{loopParEvalN}}. Finally we unzip the output list \ensuremath{[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} list into \ensuremath{([\mskip1.5mu \Varid{o}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])}.

@@

As an example of using this skeleton, \citet{Eden:SkeletonBookChapter02} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}). An adapted version can be found in Fig.~\ref{fig:torusMatMult}.

@@

@@

\begin{figure}[tb]
	\centering
	\includegraphics[width=0.9\textwidth]{torus_matrix_eden_scale}
	\caption[Matrix multiplication with \ensuremath{\Varid{torus}} (Eden)]{Matrix multiplication with \ensuremath{\Varid{torus}} (Eden).}
	\label{fig:torus_eden_trace}
\end{figure}

@@

The preceding section has shown that PArrows are expressive. This section evaluates the performance overhead of this compositional abstraction in comparison to  GpH and the \ensuremath{\Conid{Par}} Monad on shared memory architectures and Eden on  a distributed memory cluster.
We describe our measurement platform, the benchmark results -- the shared-memory variants (GpH, \ensuremath{\Conid{Par}} Monad and Eden~CP) followed by Eden in a distributed memory setting, and conclude that PArrows hold up in terms of performance when compared to the original parallel Haskells.

>>>>

We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner---we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel implementations at will. The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}). We have also conducted some initial experiments with HdpH and Frege; those experiments inducate that our parallel Arrows DSL can be relatively easily extended to further languages.

@@

\ptcomment{The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations. A range of approaches have been explored, including data parallelism (Chakravarty et al., 2007; Keller et al.,
2010), GPU-based approaches (Mainland & Morrisett, 2010; Svensson, 2011), software
transactional memory (Harris et al., 2005; Perfumo et al., 2008). The Haskell–GPU bridge
Accelerate (Chakravarty et al., 2011; Clifton-Everest et al., 2014; McDonell et al., 2015)
is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.

@@

Other task parallel Haskells, related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad, include the following.  
HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of \ensuremath{\Conid{Par}} Monad to
heterogeneous clusters. LVish (Kuper et al., 2014) is a communication-centred extension
of \ensuremath{\Conid{Par}} Monad.}
We chose to not extend our implementation to LVish and HdpH at the moment. The former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper. The latter heavy relies on Template Haskell, which adds complexity to the implementation.

@@

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, in essence they are a generalised function arrow~\ensuremath{\to }. \citet{Hughes2005} presents a tutorial on Arrows. \citet{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} develop theoretical background of Arrows. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular Arrows do and thus enable optimisations. Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of a list of Arrows. \citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.

\paragraph{Arrows in other languages.}
Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined. Preliminary experiments show that basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM). However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}}, 
a new approach to Haskell on the JVM.

@@

While we could have based our DSL on Profunctors \cite{boisseau_what_2018} as well, we chose Arrows\mbcomment{cite missing?} for this paper since they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability. However, they are a promising candidate for future improvements of our DSL. Some Profunctors, especially ones supporting a composition operation, choice, and looping, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
	\subsection{Short introduction to parallel Haskells}
\label{sec:parallelHaskells}
\label{sec:parEvalNIntro}
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel or \ensuremath{\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{parEvalN}
	\caption{Schematic illustration of \ensuremath{\Varid{parEvalN}}. A list of inputs is transformed by different functions in parallel.}
	\label{fig:parEvalN}
\end{figure}

In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow-based parallel Haskell.
As the goal of this paper is not to re-implement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface. For this paper we chose GpH for its simplicity, the \ensuremath{\Conid{Par}} Monad to represent a monadic DSL, and Eden as a distributed parallel Haskell.

@@

\paragraph{Eden TraceViewer.}
\label{sec:edentv}
To comprehend the efficiency or the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}. In the next sections we will present some \emph{trace visualisations}, the post-mortem process diagrams of Eden processes and their activity.

@@

\subsubsection{\ensuremath{\Conid{Par}} Monad}
\olcomment{introduce a newcommand for par-monad, "Arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
As for GpH we can easily lift the definition of \ensuremath{\Varid{parEvalN}} for the \ensuremath{\Conid{Par}} Monad to Arrows in Fig.~\ref{fig:ArrowParallelParMonad}. To start off, we define the \ensuremath{\Conid{Strategy}\;\Varid{a}} and \ensuremath{\Conid{Conf}\;\Varid{a}} type so we can have a configurable instance of ArrowParallel:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{Strategy}\;\Varid{a}\mathrel{=}\Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{Conf}\;\Varid{a}\mathrel{=}\Conid{Conf}\;(\Conid{Strategy}\;\Varid{a}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Now we can once again define our \ensuremath{\Conid{ArrowParallel}} instance as follows: First, we convert our Arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{evalN}\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{strat})\;\Varid{fs})} into an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu (\Conid{Par}\;(\Conid{IVar}\;\Varid{b}))\mskip1.5mu]} that yields composable computations in the \ensuremath{\Conid{Par}} monad. By combining the result of this Arrow with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}, we get an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu])}. Then, in order to fetch the results of the different threads, we map over the \ensuremath{\Conid{IVar}}s inside the \ensuremath{\Conid{Par}} Monad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}---our intermediary Arrow is of type \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu])}. Finally, we execute the computation \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} by composing with \ensuremath{\Varid{arr}\;\Varid{runPar}} and get the final Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{parEvalN}\;(\Conid{Conf}\;\Varid{strat})\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{evalN}\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{strat})\;\Varid{fs})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{\Conid{Control}.\Conid{Monad}.\Conid{Par}.get})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{runPar}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{\ensuremath{\Conid{Par}} Monad \ensuremath{\Conid{ArrowParallel}} instance.}
\label{fig:ArrowParallelParMonad}
\end{figure}

@@

\subsubsection{Default configuration instances}
While the configurability in the instances of the \ensuremath{\Conid{ArrowParallel}} interface above is nice, users probably would like to have proper default configurations for many parallel programs as well. These can also easily be defined as we can see by the example of the default implementation of \ensuremath{\Conid{ArrowParallel}} for the GpH:

@@

This is only a problem in distributed memory (in the scope of this paper) and we should allow nodes to communicate directly with each other. Eden already provides \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
But as we want code with our DSL to be portable and implementation agnostic, we have to wrap this concept. We do this with the \ensuremath{\Conid{Future}} type class (Fig.~\ref{fig:future}). We require a \ensuremath{\Varid{conf}} parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the \ensuremath{\Conid{ArrowParallel}} type class earlier.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{Future}\;\Varid{fut}\;\Varid{a}\;\Varid{conf}\mid \Varid{a}\;\Varid{conf}\to \Varid{fut}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{put}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{conf}\to \Varid{arr}\;\Varid{a}\;(\Varid{fut}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{get}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{conf}\to \Varid{arr}\;(\Varid{fut}\;\Varid{a})\;\Varid{a}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The \ensuremath{\Conid{Future}} type class.}
\label{fig:future}
\end{figure}
Since \ensuremath{\Conid{RD}} is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows. %This is due to the same reason we had to introduce a wrapper for |Strategy a| in the GpH Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
Technical details are in Appendix, in Section~\ref{app:omitted}.

@@

Next, we zip the resulting \ensuremath{([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])} to \ensuremath{[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})}. We then feed this into our parallel Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]\;[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} obtained by transforming our input Arrow \ensuremath{\Varid{f}\mathbin{::}\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})} into \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})} before \ensuremath{\Varid{repeat}}'ing and lifting it with \ensuremath{\Varid{loopParEvalN}}. Finally we unzip the output list \ensuremath{[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} list into \ensuremath{([\mskip1.5mu \Varid{o}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])}.

@@

As an example of using this skeleton, \citet{Eden:SkeletonBookChapter02} showed the matrix multiplication using Gentleman's algorithm (\citeyear{Gentleman1978}). An adapted version can be found in Fig.~\ref{fig:torusMatMult}.

@@

%
% \begin{figure}[tb]
	\includegraphics[width=0.9\textwidth]{torus_matrix_eden_scale}
	\caption[Matrix multiplication with \ensuremath{\Varid{torus}} (Eden)]{Matrix multiplication with \ensuremath{\Varid{torus}} (Eden).}
	\label{fig:torus_eden_trace}

@@

@@

The preceding section has shown that PArrows are expressive. This section evaluates the performance overhead of this compositional abstraction in comparison to  GpH and the \ensuremath{\Conid{Par}} Monad on shared memory architectures and Eden on  a distributed memory cluster.
We describe our measurement platform, the benchmark results---the shared-memory variants (GpH, \ensuremath{\Conid{Par}} Monad and Eden~CP) followed by Eden in a distributed memory setting, and conclude that PArrows hold up in terms of performance when compared to the original parallel Haskells.
